Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models

### 2018-08-21 paper.tex
Update on Overleaf.

% **************  TITLE AND AUTHOR INFORMATION **************

@@

\title{Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models
%Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?
}

@@

\address[a]{UPS CNRS 3611 ISC-PIF, Paris, France}
\address[b]{UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Paris, France}
\address[c]{Centre for Advanced Spatial Analysis, University College London, UK}
\address[d]{UMR CNRS 8097 Centre Maurice Halbwachs, Paris, France}
\address[e]{UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France}
\address[f]{Université Paris-Est, Laboratoire Ville Mobilité Transport, Marne-la-Vallée, France}
%NB : mon affiliation "officielle" c'est : 
%Le Néchet, F., Université Paris-Est, Laboratoire Ville Mobilité Transport, UPEMLV : 5 boulevard Copernic, Cité Descartes F 77454 Marne-la-Vallée cedex 2 France

@@

The last methodological point which we need to emphasize is the relation between the present workflow and model exploration workflows in general. The ideas of multi-modeling and extensive model exploration are nothing from new as ~\cite{openshaw1983data} already advocated for ``model-crunching'', but their effective use only begins to emerge thanks to the development of new methods and tools together with an explosion of computation capabilities. The model exploration platform OpenMOLE~\citep{reuillon2013openmole} allows to embed any model as a blackbox, to write modulable exploration workflow using advanced methodologies such as genetic algorithms and to distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids. In our case, this tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and to allow the coupling of any spatial generator with any model in a straightforward way as long as the model can take its spatial configuration as an input or from an input file. In this paper, we use the OpenMOLE platform for the spatial environment and the model coupling, placing ourselves in the  framework of multi-modelling \citep{cottineau2015modular}.

@@

%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[htbp] \begin{center} 
\resizebox{0.9\textwidth}{!}{ 
	\includegraphics{figures/spatialGen.png}
} \caption{Four examples of grids produced by the spatial generator. The lighter the red, the denser the area. Changing the growth rate $N$ allows to have more or less chaotic shapes (two first compared to the two last grids for example) corresponding to different levels of convergence of the model, whereas local radius can be tuned with the interplay of aggregation strength $\alpha$ and diffusion strength $\beta$.} \label{fig:spatialGen} \end{center} \end{figure*} %
%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/relativedistance_metaparams}\\
%\includegraphics[width=0.4\textwidth]{figures/alpha_beta_schemas}(Bottom) Corresponding stylised density grids\\\comment[JR]{je sais pas si c'est une bonne idee de mettre les grilles stylisees comme ca parce que ca donne une fausse idee des sorties du modele de generation}}
\caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by the meta-parameters of their generator.} Relative distance as a function of meta-parameters $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).}
\label{fig:sugarscape-distance-meta}
\end{figure}
%%%%%%%%%%%%%

@@

After reviewing the extensive literature on spatial biases in statistical and simulation models, we presented a method to analyse the sensitivity of a simulation's results to the initial spatial configuration. We did so by implementing a spatial generator whose output is used as input for the simulation model. We applied this approach to two mythical ABMs: Schelling and Sugarscape. With the Schelling experiment, we found that the different urban morphologies impact the interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index. With Sugarscape, we show that the model is more sensitive to space than to its other parameters in the reference Netlogo implementation, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behaviour of the phase diagram is qualitatively different in different regions of the morphological space. We think that this method has the potential to increase the arsenal of evaluation of geographical models, in order to assess the sensitivity of models to their initial spatial conditions but also to learn about the impact of the urban form on social mechanisms.
%%%%%%%%%%%%%
% Acknowledgements

% acks commentés pour la soumission (anonymous)
%\begin{acks}
%The authors acknowledge the funding of their institutions and the EPSRC project number EP/M023583/1. Results obtained in this paper were computed on the vo.complex-system.eu virtual organization of the European Grid Infrastructure ( http://www.egi.eu ). We thank the European Grid Infrastructure and its supporting National Grid Initiatives (France-Grilles in particular) for providing the technical support and infrastructure.
%\end{acks}

@@

@@

>>>>

@@

\title{Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models}

@@

\address[a]{UPS CNRS 3611 ISC-PIF, Paris, France}
\address[b]{UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Paris, France}
\address[c]{Centre for Advanced Spatial Analysis, University College London, UK}
\address[d]{UMR CNRS 8097 Centre Maurice Halbwachs, Paris, France}
\address[e]{UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France}
\address[f]{Université Paris-Est, Laboratoire Ville Mobilité Transport, Marne-la-Vallée, France}

@@

The last methodological point which we need to emphasize is the relation between the present workflow and model exploration workflows in general. The ideas of multi-modeling and extensive model exploration are nothing from new as ~\cite{openshaw1983data} already advocated for ``model-crunching'', but their effective use only begins to emerge thanks to the development of new methods and tools together with an explosion of computation capabilities. The model exploration platform OpenMOLE~\citep{reuillon2013openmole} allows to embed any model as a blackbox, to write modulable exploration workflow using advanced methodologies such as genetic algorithms and to distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids. In our case, this tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and to allow the coupling of any spatial generator with any model in a straightforward way as long as the model can take its spatial configuration as an input or from an input file. In this paper, we use the OpenMOLE platform for the spatial environment and the model coupling, placing ourselves in the  framework of multi-modelling \citep{cottineau2015modular}. \added{We use therefore OpenMOLE's functionalities for model embedding through workflow, design of experiments (parameter sampling) and high performance environment access.}

@@

%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[htbp] \begin{center} 
	\includegraphics[width=\linewidth]{figures/spatialGen.png}
 \caption{\textbf{Four examples of grids produced by the spatial generator.} The lighter the red, the denser the area. Changing the growth rate $N$ allows to have more or less chaotic shapes (two first compared to the two last grids for example) corresponding to different levels of convergence of the model, whereas local radius can be tuned with the interplay of aggregation strength $\alpha$ and diffusion strength $\beta$. \added{Parameter values used here are for the first grid $(\alpha = 0.4,\beta=0.006, N = 25, n_d = 1, t_f = 971)$, the second $(\alpha = 0.4,\beta=0.006, N = 25, n_d = 1, t_f = 176)$, the third $(\alpha = 1.4,\beta=0.045, N = 102, n_d = 2, t_f = 618)$, and the fourth $(\alpha = 1.8,\beta=0.114, N = 108, n_d = 1, t_f = 227)$.}}
\label{fig:spatialGen}
\end{center}
\end{figure*} %
%%%%%%%%%%%%%%%%%%%%%%

% Generator parameters values:
%  1) ex_sp-diffusion=0.0060_sp-growth-rate=25_sp-diffusion-steps=1_sp-alpha-localization=0.4_ticks=971_sp-population=24275.000000000102
%  2) ex_sp-diffusion=0.0060_sp-growth-rate=25_sp-diffusion-steps=1_sp-alpha-localization=0.4_ticks=176_sp-population=4400.000000000003
%  3) ex_sp-max-pop=80260_sp-diffusion=0.045_sp-growth-rate=102_sp-diffusion-steps=2_sp-alpha-localization=1.4_ticks=618
%  4) ex_sp-max-pop=382170_sp-diffusion=0.114_sp-growth-rate=108_sp-diffusion-steps=1_sp-alpha-localization=1.8_ticks=227

@@

%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/relativedistance_metaparams}
\caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by the meta-parameters of their generator.} Relative distance as a function of meta-parameters $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).}
\label{fig:sugarscape-distance-meta}
\end{figure}
%%%%%%%%%%%%%

@@

After reviewing the extensive literature on spatial biases in statistical and simulation models, we presented a method to analyse the sensitivity of a simulation's results to the initial spatial configuration. We did so by implementing a spatial generator whose output is used as input for the simulation model. We applied this approach to two \replaced{well-known}{mythical} ABMs: Schelling and Sugarscape. With the Schelling experiment, we found that the different urban morphologies impact the interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index. With Sugarscape, we show that the model is more sensitive to space than to its other parameters in the reference Netlogo implementation, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behaviour of the phase diagram is qualitatively different in different regions of the morphological space. We think that this method has the potential to increase the arsenal of evaluation of geographical models, in order to assess the sensitivity of models to their initial spatial conditions but also to learn about the impact of the urban form on social mechanisms.
%%%%%%%%%%%%%
% Acknowledgements

%\begin{acks}
%The authors acknowledge the funding of their institutions and the EPSRC project number EP/M023583/1. Results obtained in this paper were computed on the vo.complex-system.eu virtual organization of the European Grid Infrastructure ( http://www.egi.eu ). We thank the European Grid Infrastructure and its supporting National Grid Initiatives (France-Grilles in particular) for providing the technical support and infrastructure.
%\end{acks}

@@

\subsection*{\added{Behavior of the density grid generator}}

\added{We give in Fig.~\ref{fig:stylized-generator} plots roughly summarizing the behavior of the density grid generator, according to parameters $(\alpha,\beta)$.}

%%%%%%%%%%%%%%
\begin{figure}
	\includegraphics[width=0.49\linewidth]{figures/alpha_beta_examples.png}
    \includegraphics[width=0.49\linewidth]{figures/alpha_beta_schemas.png}
    \caption{\added{\textbf{Stylized behavior of the density grid generator.}}\label{fig:stylized-generator}}
\end{figure}
%%%%%%%%%%%%%%

@@

% Minimal netlogo code for the sugarscape model
\begin{lstlisting}
	to main-loop
    	regrow-patches
        update-agetns
    end

    to regrow-patches
    	ask patches [set psugar min (list max-psugar (psugar + 1))]
    end

    to update-agents
    	ask turtles [
    		let move-candidates (patch-set patch-here (patches at-points vision-points) with [not any? turtles-here])
  let possible-winners move-candidates with-max [psugar]
  if any? possible-winners [
    ;; if there are any such patches move to one of the patches that is closest
    move-to min-one-of possible-winners [distance myself]
  ]

            ; 
            set sugar (sugar - metabolism + psugar)
  			set psugar 0

    		set age (age + 1)
    		if sugar <= 0 or age > max-age [
      			hatch 1 [ new-turtle ]
      			die
    		]
  		]
    end

\end{lstlisting}
